---
title: "HW2"
author: "Karthik Bobba"
date: "2025-02-11"
output: html_document
---

```{r loading libraries and data}
library(tidyverse)
library(lubridate)
library(kableExtra)
#loading neccessary libraries

inspections <- read_csv("restaurant_inspections.csv") #loading data set into a df called inspections

inspections <- inspections %>%
  mutate(
    # Convert date columns to proper datetime format
    DATE_ = ymd_hms(DATE_),
    RESTAURANTOPENDATE = ymd_hms(RESTAURANTOPENDATE),
    
    # Standardize city names for question 3
    CITY = str_to_upper(CITY) %>% 
      recode("RTP" = "RESEARCH TRIANGLE PARK",
             "CARY" = "CARY",
             "Apex" = "APEX",
             "RALEIGH" = "RALEIGH",
             "FUQUAY-VARINA" = "FUQUAY VARINA",
             "HOLLY SPRING" = "HOLLY SPRINGS",
             "MORRISVILE" = "MORRISVILLE",
              ) 
  )
```

# 1) Histogram

```{r histogram}
ggplot(inspections, aes(SCORE)) +
  geom_histogram(binwidth = 2, fill = "blue", color = "white") +
  labs(title = "Distribution of Inspection Scores",
       x = "Score", y = "Count")

#essentially creating a histogram with score as the x axis and count as the y axis, showing the distribution of scores
```

# 2) Trend for New vs Old Facility

We don't really see any trend visually

```{r old vs new}
inspections %>%
  mutate(restaurant_age = time_length(DATE_ - RESTAURANTOPENDATE, "years")) %>% #Essentially what is happening here is calculating how long it's been since they've opened, using the time_length function
  ggplot(aes(restaurant_age, SCORE)) +
  geom_point(alpha = 0.5) +       # Add transparency for overlapping points
  labs(title = "Inspection Scores by Restaurant Age",
       x = "Years Since Opening", y = "Score")
#We don't really see any trend visually
```

# 3) Vary by City?

Average score seems to have very minimal variation between cities

```{r by city}
city_scores <- inspections %>%
  group_by(CITY) %>%
  summarise(
    avg_score = mean(SCORE, na.rm = TRUE),
    n_inspections = n(),  # Count number of inspections per city
    .groups = "drop"
  )

# Create formatted table with kableExtra
kable(city_scores, caption = "Average Scores by City") %>%
  kable_styling()

#Average score seems to have very minimal variation between cities
```

# 4) Inspector Ratings

Average score does seem to have some variation depending on the inpsector

```{r inspectors}
# Calculate inspector performance metrics
inspector_scores <- inspections %>%
  group_by(INSPECTOR) %>%
  summarise(
    avg_score = mean(SCORE, na.rm = TRUE),
    n_inspections = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(avg_score))  # arranging by highest avg score
#Essentially finding highest avg scores by inspector
kable(head(inspector_scores, 200), caption = "Inspectors by Average Score") %>%
  kable_styling()


```

# 5) Small Samples Sizes?

Yes, in both the inspectors and the city tables we see instances where the sample size is less than 5, as you can see in the n_inspections column, which could lead to possible extreme values, skewing the data.

# 6) Restaurants vs Other

As we see in the table, the scores for restaurant (where is_restaurant is TRUE) is slightly less than for non-restaurants

```{r resaurants vs other}
inspections %>%
  mutate(is_restaurant = FACILITYTYPE == "Restaurant") %>%
  group_by(is_restaurant) %>%
  summarise(
    avg_score = mean(SCORE, na.rm = TRUE),
    n = n(),  # Total inspections in each group
    .groups = "drop"
  ) %>%
  kable(caption = "Scores by Facility Type") %>%
  kable_styling()
```

# 7) Restaurant Specific Analysis

Here we will be redoing all the prior analysis' but with a df that contains only restaurants

```{r specific analysis}
rest_only <- inspections %>%
  filter(FACILITYTYPE == "Restaurant") #Creates a df with only restaurants

#histogram
ggplot(rest_only, aes(SCORE)) +
  geom_histogram(binwidth = 2, fill = "blue", color = "white") +
  labs(title = "Distribution of Inspection Scores",
       x = "Score", y = "Count")

#new vs old
rest_only %>%
  mutate(restaurant_age = time_length(DATE_ - RESTAURANTOPENDATE, "years")) %>% #Essentially what is happening here is calculating how long it's been since they've opened, using the time_length function
  ggplot(aes(restaurant_age, SCORE)) +
  geom_point(alpha = 0.5) +       # Add transparency for overlapping points
  labs(title = "Inspection Scores by Restaurant Age",
       x = "Years Since Opening", y = "Score")

#by city
rest_city_scores <- rest_only %>%
  group_by(CITY) %>%
  summarise(
    avg_score = mean(SCORE, na.rm = TRUE),
    n_inspections = n(),  # Count number of inspections per city
    .groups = "drop"
  )

# Create formatted table with kableExtra
kable(rest_city_scores, caption = "Average Scores by City") %>%
  kable_styling()

#inspectors
# Calculate inspector performance metrics
rest_inspector_scores <- rest_only %>%
  group_by(INSPECTOR) %>%
  summarise(
    avg_score = mean(SCORE, na.rm = TRUE),
    n_inspections = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(avg_score))  # arranging by highest avg score
#Essentially finding highest avg scores by inspector
kable(head(rest_inspector_scores, 200), caption = "Inspectors by Average Score") %>%
  kable_styling()

#5) Again, sample size where it's less than 5 could yield extreme results where it can skew the overall data
```
